---
title: "Plotting for metric learning"
author: "Fan Cheng"
date: "2021-06-10"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=TRUE, results='hide'}
library(tidyverse)
library(dimRed)
library(reticulate)
library(here)
library(viridis)
library(hdrcde)
library(igraph)
library(matrixcalc)
library(akima)
library(car)
Jmisc::sourceAll(here::here("R/sources"))
set.seed(123)
```

# Metric learning

## Smart meter data for one household

```{r data}
# load(here::here("data/spdemand_3639id336tow.rda"))
# nid <- 1
# ntow <- 336
# 
# train <- spdemand %>%
#   lazy_dt() %>%
#   filter(tow <= ntow,
#          # id <= sort(unique(spdemand[,id]))[nid],
#          id == 1003) %>%
#   dplyr::select(-id, -tow) %>%
#   data.table::as.data.table()
train <- readRDS(here::here("data/spdemand_1id336tow_train.rds"))
```

## Radius searching with k-d trees

```{r parameters}
# Parameters fixed
x <- train
N <- nrow(train)
s <- 2
k <- 20
method <- "annIsomap"
annmethod <- "kdtree"
distance <- "euclidean"
treetype <- "kd"
searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
radius <- .4 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm
```

## The metric learning algorithm

```{r, message=FALSE}
metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, invert.h = TRUE)
summary(metric_isomap)
```

The function `metricML()` returns a list of

-   the embedding coordinates `embedding` of $N \times s$
-   the embedding metric `rmetric` for each point $p \in x$ as an array. By default, `invert.h = TRUE` returns the inverse of Riemannian metric. 
-   the weighted neighborhood graph as an `igraph` object `weighted_graph`
-   the sparse adjacency matrix from the graph `adj_matrix`
-   the graph laplacian matrix `laplacian`


# Embedding plot

```{r embedplot, eval=FALSE}
# Embedding plot of metricML() output for electricity data
plot.embedding <- function(x) {
  fn <- x$embedding
  
  todcolor <- colorspace::scale_color_continuous_sequential(
    palette = "viridis",
    breaks = c(0, 12, 24, 36, 48),
    labels=c("00:00", "06:00", "12:00", "18:00", "24:00"),
    name="Time of day",
    guide=guide_colorbar(barwidth = 10))
  
  p <- fn %>% 
    cbind(tod = rep(1:48, N / 48)) %>% 
    as_tibble() %>% 
    ggplot(aes(x = E1, y = E2, col = tod)) + 
    geom_point() + 
    # labs(x = "ISO1", y = "ISO2") +
    coord_fixed(ratio = 1) + 
    todcolor + 
    theme(legend.position = 'bottom')
  
  return(p)
}
```

```{r, eval=T, include=T}
# fn <- metric_isomap$embedding
# head(fn)
# plot(fn, col = viridis::viridis(24), asp = 1)
plot.embedding(metric_isomap) +
  labs(x = "ISO1", y = "ISO2")
```



# Ellipse plot

Using the Riemannian metric for each point as the 2-d covariance matrix and the 2-d embeddings as the centroid to plot an ellipse for each point.

The underlying requirement is that the Riemannian metric needs to be a square positive definite matrix.

```{r}
# Done: add argument `n.plot` to specify how many ellipse to be added
# If add = FALSE, fix xlim, ylim, radius for generating ellipse
plot.ellipse <- function(x, n.plot = 50) {
  fn <- x$embedding
  Rn <- x$rmetric
  cols <- viridis::viridis(24) 
  n.plot <- 50
  plot(fn, col = cols, asp = 1, pch = 20)
  for (i in sample(1:N, n.plot)) {
    mat <- Rn[,,i] 
    center <- c(fn[i,1], fn[i,2])
    add <- T 
    car::ellipse(center, mat, 
                 radius = .1, # controls the size of ellipse, currently set manually
                 # xlim = c(-.2, .25), ylim = c(-.2, .2),
                 # col = cols[i],
                 col = blues9[5],
                 asp = 1, pty = "s", lwd = 1, center.pch = 19, center.cex = 0,
                 fill = T, fill.alpha = 0.2, grid = T, add = add, 
                 )
  }
}
plot.ellipse(metric_isomap)
```

To plot as an ggplot2 object, the exsiting `stat_ellipse` does not work for variable covariance matrix. I tried to modify the code and create `stat_ellipse_cov()` which takes another aesthetic `covmat`. But all points are connected as one path. 

Also, `purrr:::pmap_dfc()` does not work correctly, so for loop is currently in use. 

```{r}
fn <- metric_isomap$embedding
Rn <- metric_isomap$rmetric
a <- fn %>% 
  as_tibble() %>% 
  add_column(covmat = purrr::array_branch(Rn, 3))
a
# a$covmat %>% head()

a %>% 
  cbind(tod = rep(1:48, N / 48)) %>% 
  ggplot(aes(E1, E2)) + 
  geom_point() +
  stat_ellipse()
  # stat_ellipse(aes(group = tod, color = tod))

# a %>% 
#   cbind(tod = rep(1:48, N / 48)) %>%
#   # group_by(tod) %>%
#   ggplot(aes(x = E1, y = E2, color = tod)) +
#   geom_point() +
#   scale_color_viridis() + 
#   stat_ellipse_cov(aes(covmat = covmat))

plot.embedding(x = fn, embedding = T) + 
  stat_ellipse_cov(data = cbind(a, tod = rep(1:48, N / 48)), aes(covmat = covmat))
```


# Contour plot

## Variable kernel density estimate

For multivariate data, the variable kernel density estimate is given by $$\hat{f}(x) = n^{-1} \sum_i K_{H_i} (x - X_i).$$

### Inverse of Riemannian metric as bandwidth matrix

If we use the embedding `x` and the inverse of Riemannian metric `h`, we could produce contour plots based on the density estimates.

```{r pydensity, eval=T}
emb_isomap <- feather::read_feather(here::here("data/embedding_isomap_1id336tow.feather"))
np <- reticulate::import("numpy")
H_isomap <- np$load(here::here("data/hmatrix_isomap_1id336tow.npy"))
pyh_isomap <- array(NA, dim = c(s, s, N))
for(i in 1:N){
  pyh_isomap[,,i] <- H_isomap[i,,]
}

x <- cbind(emb_isomap$`0`, emb_isomap$`1`)
f <- mkde(x = x, h = pyh_isomap)
summary(f)
# Top 20 index with the lowest densities
head(order(f), n = 20)
```

```{r, echo=FALSE, message=FALSE}
p_pyh <- plot.contour(x, f, plot.hdr = F)
```

Try linear interpolation using `approx()`.

```{r, eval=FALSE, echo=TRUE}
x_interp <- approx(x[,1], x[,2], n = 100, yleft = 0, yright = 1, ties = mean)
x_int <- cbind(x_interp$x, x_interp$y)
dim(x_int)
f_linerinterp <- mkde(x = x_int, h = pyh_isomap)
f_linerinterp
plot.contour(x_interp, f_linerinterp)
hdrscatterplot(x_interp$x, x_interp$y, noutliers = 10)
p <- filled.contour(akima.spl, color.palette = viridis,
                 plot.axes = { axis(1); axis(2);
                   # title("smooth interp(*, linear = FALSE)");
                   points(embed_den, pch = 3, col= hcl(c=20, l = 10))})
```

Now apply the `mkde()` function to the output of `metricML()` written in R.

```{r rdensity}
x <- metric_isomap$embedding
riem_isomap <- metric_isomap$rmetric
f <- mkde(x = x, h = riem_isomap)
summary(f)
head(order(f), n=20)
```

```{r, echo=FALSE, message=FALSE}
p_rh <- plot.contour(x, f, plot.hdr = F)
```

We can check the outliers based on density `f`.
In R, the outliers are similar to the `megaman` output.


<!-- ### Riemannian metric itself as bandwidth matrix -->

<!-- ```{r pydensity1, eval=T} -->
<!-- R_isomap <- np$load(here::here("data/rmatrix_isomap_1id336tow.npy")) -->
<!-- # R_isomap[1,,, drop=TRUE] %>%  -->
<!-- #   matrixcalc::is.positive.definite() # not necessarily symmetric -->
<!-- pyr_isomap <- array(NA, dim = c(s, s, N)) -->
<!-- for(i in 1:N){ -->
<!--   pyr_isomap[,,i] <- R_isomap[i,,] -->
<!-- } -->
<!-- pyr_isomap[,,1:3] -->

<!-- x <- cbind(emb_isomap$`0`, emb_isomap$`1`) -->
<!-- f <- mkde(x = x, h = pyr_isomap) -->
<!-- summary(f) -->
<!-- # Top 20 index with the lowest densities -->
<!-- head(order(f), n=20) -->
<!-- p_pyr <- plot.contour(x, f, plot.hdr = F) -->
<!-- ``` -->

<!-- Now apply the `mkde()` function to the output of `metricML()` written in R. -->

<!-- ```{r rdensity1} -->
<!-- hn <- riemann_metric(Y = as.matrix(emb_isomap), laplacian = Ln, d = s, invert.h = T) -->
<!-- f <- mkde(x = metric_isomap$embedding, h = hn) -->
<!-- summary(f) -->
<!-- head(order(f), n=20) -->
<!-- p_rr <- plot.contour(x, f, plot.hdr = F) -->
<!-- ``` -->



# t-SNE

```{r tsne, message=FALSE, warning=FALSE, eval=TRUE}
method <- "tSNE"
perplexity <- 30
theta <- 0.5
metric_tsne <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, perplexity = perplexity, theta = theta, invert.h = TRUE)
summary(metric_tsne)
```

```{r, include=FALSE, eval=FALSE}
perplexity <- 30
theta <- 0.5
method <- "tSNE"
e <- dimRed::embed(x,
                     .method = method,
                     knn = k,
                     ndim = s,
                     annmethod = annmethod,
                     radius = radius,
                     eps = eps,
                     nt = nt,
                     nlinks = nlinks,
                     ef.construction = ef.construction,
                     distance = distance,
                     treetype = treetype,
                     searchtype = searchtype,
                     perplexity = perplexity,
                     theta = theta,
                     .mute = c("output"),
                     # ...

  )
  fn <- e@data@data
plot(metric_tsne$embedding)
```

```{r, message=FALSE, eval=TRUE}
f <- mkde(x = metric_tsne$embedding, h = metric_tsne$rmetric)
summary(f)
head(order(f), n=20)
plot.contour(x = metric_tsne$embedding, f, plot.hdr = F)
```


```{r anntsne, message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}
x <- train
nn2res <- dplyr::case_when(distance=="euclidean" ~ 
                                 RANN::nn2(data = x, query = x, k = k + 1, 
                                           treetype = treetype, searchtype = searchtype, eps = eps,
                                           radius = radius),
                               distance=="manhattan" ~ 
                                 RANN.L1::nn2(data = x, query = x, k = k + 1, 
                                              treetype = treetype, searchtype = searchtype, eps = eps,
                                              radius = radius),
    )
    names(nn2res) <- c("nn.idx", "nn.dists")
    
metric_tsne1 <- metricML(x, nn.idx = nn2res$nn.idx, nn.dists = nn2res$nn.dists, s = s, k = k, radius = radius, method = "anntSNE", annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, perplexity = k-1, theta = 0.5, invert.h = TRUE)
summary(metric_tsne1)


f <- mkde(x = metric_tsne1$embedding, h = metric_tsne1$rmetric)
summary(f)
head(order(f), n=20)
plot.contour(x = metric_tsne1$embedding, f, plot.hdr = F)
```







```{r, echo=FALSE, eval=FALSE}
# Countour plot for megaman, dual metric
# x <- cbind(emb_isomap$`0`, emb_isomap$`1`)
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))})
```

```{r, echo=FALSE, include=FALSE, eval=FALSE}
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
# library(akima)
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
p.zmin <- min(embed_den$z,na.rm=TRUE)
p.zmax <- max(embed_den$z,na.rm=TRUE)
breaks <- pretty(c(p.zmin,p.zmax), lenbreak)
# hdrcde
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
hdrinfo <- hdr.2d(embed_den$x, embed_den$y, prob = c(50, 95, 99))
plot.hdr2d(hdrinfo)
# Contour
contour(akima.spl, main = "smooth interp(*, linear = FALSE)", col = viridis(length(breaks)-1), levels=breaks, add=TRUE)
points(embed_den, pch = 20)

filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))},
               # plot.title = {par(cex.main=1);title(main = "The Topography of Maunga Whau", xlab = "Meters North", ylab = "Meters West")},
               # plot.axes = { axis(1, seq(100, 800, by = 100))
                 # axis(2, seq(100, 600, by = 100)) },
               # key.title = {par(cex.main=0.3);title(main="Height\n(meters)")},
               # key.axes = axis(4, seq(-1, 1.5, by = 0.5), asp=1)
)

```
